---
title: "NBA Data Description"
author: "Ben Moolman, Craig Orman, Ethan Pross"
output: pdf_document
---

```{r, warning = FALSE, message = FALSE, echo = FALSE}
library(dplyr)
library(tidyverse)
library(knitr)
library(skimr)

original_tbl <- read.csv("../NBA-BoxScores-2023-2024.csv")
# original_tbl$COMMENT[original_tbl$COMMENT == ""] <- NA
# original_tbl$COMMENT <- factor(original_tbl$COMMENT)

original_tbl <- mutate(original_tbl,
                       START_POSITION = na_if(START_POSITION, ""),
                       START_POSITION = factor(original_tbl$START_POSITION),
                       COMMENT = na_if(COMMENT, ""),
                       COMMENT = factor(original_tbl$COMMENT),
                       MIN = na_if(MIN, ""),
                       )

original_tbl = original_tbl[original_tbl$START_POSITION != "",]

# Unknown variables requiring decoding
# -min

```

## Data Set Description/ Exploratory Data Analysis:

  + Obtained from the Kaggle Dataset [NBA Boxscore - Season 2023 / 2024 by Alberto Filosa](https://www.kaggle.com/datasets/albi9702/nba-boxscore-season-2023-2024?resource=download)[^1]
  
  [^1]: NBA Boxscore - Season 2023 / 2024 by Alberto Filosa, https://www.kaggle.com/datasets/albi9702/nba-boxscore-season-2023-2024?resource=download

  + There are 30 columns and `r nrow(original_tbl)` rows. 7 columns are identifiers, 1 column is a row index, 2 are characteristic and comment columns, and then we have 20 numerical statistic columns.
  
  + Each row significes a unique tuple of (game, player), that is to say there is a row for stats of each player in every game. 
  
  + There are `r length(unique(original_tbl$GAME_ID))` unique games, `r length(unique(original_tbl$TEAM_ID))` unique teams, and `r length(unique(original_tbl$PLAYER_ID))` unique players.
  
  + There is no missing data because we are only analyzing rows where the player was a starter, and therefore always played and stats were recorded.
  
  + Presumably, missing values in the starting position column refer to a non-starter player.
  
  + Most of the numeric fields are right skewed
  
  + FGA follows a binomial but with a heavier right tail
  
```{r, echo=FALSE}
df <- data.frame(
  Variable = c("Game_ID", "Team_ID", "Player_ID", "FGA", "FG_PCT", 
               "FG3A", "FG3_PCT", "REB", "AST", "PTS"),
  Description = c(
    "Unique Identifier of the Game.",
    "Unique Identifier of the Team. For Each GAME_ID there are only two TEAM_ID.",
    "Unique Identifier of the Player.",
    "Number of Field Goals (both 2 and 3 Points) Attempted by the player in the game.",
    "Percentage of Field Goals (both 2 and 3 Points) by the player in the game.",
    "Number of 3 Points Attempted by the player in the game.",
    "Percentage of 3 Points by the player in the game.",
    "Number of Total Rebounds (Defensive and Offensive Rebounds) reached by the player in the single game.",
    "Number of Assists done by the player in the single game.",
    "Number of Points done by the player in the single game."
  )
)
kable(df)
```
  
## Important notes

  + Data Analysis is on the raw data. We will be focusing on starting players.
  
```{r, echo = FALSE}
temp <- skim(original_tbl[,c(13,14,15, 16, 17, 23, 24, 29)])
colnames(temp) <- c("skim_type", "Variable", "Missing",  "complete_rate", "Mean",  "SD",  "Min", "numeric.p25",  "Median", "numeric.p75", "Max", "numeric.hist")
kable(temp[,c(2,3,5,6,9, 7, 11)])
```
## Prior Analysis on this dataset

  + There appears to be one analysis submitted for this Kaggle dataset, [Done by Adam Briggs](https://www.kaggle.com/code/adambriggs/nba-exploratory-data-analysis), they seem to have done mostly visuals and descriptive statistics, looking at best player on the team, and top performing player in various metrics. 
  
  + Basketball analytics is a very common field though, and there are entire websites dedictated to it such as one by [Evan Miyakawa](https://evanmiya.com/), who launched their website as part of a PhD program with Baylor. Usually, analytics attempt to gain an edge in the industry by identifying undervalued players or by predicting the outcomes of games.
  
## Sampling models

  + Two variables of interest to predict are Field Goal Attempts and Field Goal Percent. Given below, we will attempt to model these with the beta and binomial respectively.
  
```{r, echo=FALSE}
set.seed(45208937)
par(mfrow=c(2,2))
y1 = 2500
hist(original_tbl$FGA[original_tbl$START_POSITION != ""], breaks = 20, xlim=c(0,40), ylim = c(0, y1), main="FGA", xlab= "Field Goals Attempted")
hist(rbinom(12300, 50, 0.2), xlim=c(0,40), ylim = c(0, y1), main="Binom(50, 0.2) Approx",  xlab= "Samples")
y2 = 2500
hist(original_tbl$FG_PCT[original_tbl$START_POSITION != ""], breaks = 20, ylim=c(0, y2), main="FG_pct",  xlab= "Field Goal Percent")
hist(rbeta(12300, 7, 7), xlim=c(0,1), ylim=c(0, y2), main="Beta(7,7) Approx",  xlab= "Samples")
```

\pagebreak

# Appendix

| Variable              | Description |
|-----------------      |-------------|
| **GAME_ID**           | Unique Identifier of the Game. |
| **TEAM_ID**           | Unique Identifier of the Team. For Each GAME_ID there are only two TEAM_ID. |
| **TEAM_ABBREVIATION** | Abbreviation of the Team (e.g. GWS - Golden State Warriors). |
| **TEAM_CITY**         |  NA |
| **PLAYER_ID**         | Unique Identifier of the Player. |
| **PLAYER_NAME**       | Complete Name (Name and Surname) of the Player who played the game. |
| **NICKNAME**          | Nickname of the Player who played the game. |
| **START_POSITION**    | Position in which the player started the game (If populated, the player started the game). |
| **COMMENT**           |  NA |
| **MIN**               | Number of minutes in which the player played the game. |
| **FGM**               | Number of Field Goals (both 2 and 3 Points) Made by the player in the game. |
| **FGA**               | Number of Field Goals (both 2 and 3 Points) Attempted by the player in the game. |
| **FG_PCT**            | Percentage of Field Goals (both 2 and 3 Points) by the player in the game. |
| **FG3M**              | Number of 3 Points Made by the player in the game. |
| **FG3A**              | Number of 3 Points Attempted by the player in the game. |
| **FG3_PCT**           | Percentage of 3 Points by the player in the game. |
| **FTM**               | Number of Free Throws Made by the player in the game. |
| **FTA**               | Number of Free Throws Attempted Made by the player in the game. |
| **FT_PCT**            | Percentage of Free Throws by the player in the game. |
| **OREB**              | Number of Offensive Rebounds reached by the player in the single game. |
| **DREB**              | Number of Defensive Rebounds reached by the player in the single game. |
| **REB**               | Number of Total Rebounds (Defensive and Offensive Rebounds) reached by the player in the single game. |
| **AST**               | Number of Assists done by the player in the single game. |
| **STL**               | Number of Steals done by the player in the single game. |
| **BLK**               | Number of Blocks done by the player in the single game. |
| **TO**                | Number of Turnovers done by the player in the single game. |
| **PF**                | Number of Personal Fouls done by the player in the single game. |
| **PTS**               | Number of Points done by the player in the single game. |
| **PLUS_MINUS**        | Number of Plus Minus done by the player in the single game. |

```{r}
library(corrplot)
# Comes out as all NAs for some reason
no_na_numerics_tbl = na.omit(original_tbl[,12:30])
correlation = cor(no_na_numerics_tbl, method = "pearson")
par(mfrow=c(1,1))
corrplot(correlation) # colorful number
```